Certified Knowledge Compilation \\ with Application to Verified Model Counting

### 2023-03-19 doc/mc-paper.tex
adjust phi/theta notation

The defining clauses for a product or sum
operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
For the
extension variable $u$ associated with any POG node $\nodeu$, we can therefore
assume that any total assignment $\assign$ to the input variables will
also assign a value to $u$ such that $\assign(u) =
1$ if and only if $\assign \in \modelset(\nodeu)$.  For a POG with
root literal $r$, we can define its set of models $\modelset(P)$ as
the set of all total assignments $\assign$ such that $\assign(r) = 1$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $r$ and the defining clauses of the POG nodes. We can write
$\modelset(P)$ as a shorthand for $\modelset(\pogformula)$, recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(P)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(P) \subseteq \modelset(\inputformula)$.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root literal $r$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(P)$, i.e.,
that any total assignment $\alpha$ that is a model of $\inputformula$
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$r$ as a unit clause.  We have devised two methods of generating this
proof.  The {\em monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The {\em structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It too may require calls to a proof-generating SAT
solver.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\pogformula$.
It can
express the negation of $\pogformula$ as $\neg \pogformula = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication
will hold
when $\inputformula \imply \pogformula$, or  equivalently,
that the formula $\inputformula \land \neg \pogformula$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root literal $r$ requires showing that
$\modelset(P) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\phi$.  Deleting $C$ requires proving that
$\phi \imply C$.

@@

Each input clause deletion can be done as a single RUP step.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.

>>>>

The defining clauses for a product or sum
operation uniquely define the value of its extension variable for any assignment of values to the argument variables.
For the
extension variable $u$ associated with any POG node $\nodeu$, we can therefore
prove that any total assignment $\assign$ to the input variables which
also satisfies the POG definitions will
assign a value to $u$ such that $\assign(u) =
1$ if and only if $\assign \in \modelset(\nodeu)$.

@@

The sequence of operator declarations, asserted clauses, and
clause deletions represents a systematic transformation of the input formula
into a POG formula.  Validating all of these steps serves to prove that the
POG is logically equivalent to the input formula.
At the completion of the proof, the following conditions must hold:
\begin{enumerate}
\item There is exactly one remaining clause that was added via RUP
  addition, and this is a unit clause consisting of root literal $r$.
\item All of the input clauses have been deleted.
\end{enumerate}
In other words, at the end of the proof it must hold that the active clauses be exactly those
in $\pogformula = \{\{r\}\} \cup \; \bigcup_{\nodeu \in P} \theta_{u}$, the formula consisting
of unit clause $r$ and the defining clauses of the POG nodes. Recognizing that
any assignment to the input variables implicitly defines the assignments to the extension variables,
we can see that $\pogformula$ is the clausal encoding of $\phi_\noder$.
Let $\inputformula$ denote the input formula.
The sequence of clause addition steps provides a {\em forward implication} proof that
$\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$.  That is, any total
assignment $\assign$ satisfying the input formula must also satisfy
the formula represented by the POG\@.
Conversely,
each proof step that deletes an input clause proves that any
total assignment $\alpha$ that falsifies the clause must
assign $\assign(r) = 0$.  Deleting all but the final asserted clause and all input clauses provides a {\em reverse implication} proof
that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.

@@

For input formula $\inputformula$ and its translation into a POG $P$
with root node $\noder$, the most challenging part of the proof is to
show that $\modelset(\inputformula) \subseteq \modelset(\phi_\noder)$, i.e.,
that any total assignment $\assign$ that is a model of $\inputformula$
and the POG definition clauses
will yield $\assign(r) = 1$, for root literal $r$.  This part of the
proof consists of a series of clause assertions leading to one adding
$r$ as a unit clause.  We have devised two methods of generating this
proof.  The {\em monolithic} approach makes just one call to a
proof-generating SAT solver and has it determine the relationship
between the two representations.  The {\em structural} approach
recursively traverses the POG, generating proof obligations at each
node encountered.  It too may require calls to a proof-generating SAT
solver.

@@

The monolithic approach
takes advantage of the clausal representations of
the input formula $\inputformula$ and the POG formula $\phi_\noder$.
It can
express the negation of $\phi_\noder$ as $\neg \phi_\noder = \bigcup_{\nodeu\in P} \simplify{\theta_{u}}{\{\obar{r}\}}$.
Forward implication
will hold
when $\inputformula \imply \phi_\noder$, or  equivalently,
that the formula $\inputformula \land \neg \phi_\noder$
is unsatisfiable, where the
conjunction can be expressed as the union
of the two sets of clauses.  The proof generator writes the clauses to a file and invokes a proof-generating SAT solver.
For each clause $C$ in the unsatisfiability proof, it adds clause $\{r\} \cup C$ to the CRAT proof, and so the empty clause in the proof becomes the unit clause $r$.
Our experimental results show
that this approach can be very effective and generates short proofs
for smaller problems, but it does not scale well enough for routine
use.

@@

Completing the equivalence proof of input formula $\inputformula$ and its POG
representation with root node $\noder$ requires showing that
$\modelset(\phi_\noder) \subseteq \modelset(\inputformula)$.  This is done in the
CRAT framework by first deleting all asserted clauses, except for the
final unit clause for root literal $r$, and then deleting all of the
input clauses.  Suppose at some point in this process that we want to
delete clause $C$ such that the remaining set of clauses will form the
CNF formula $\phi$.  Deleting $C$ requires proving that
$\phi \imply C$.

@@

Each input clause deletion can be done as a single RUP step.  The
proof generator constructs the hint sequence from the defining
clauses of the POG nodes via a single, bottom-up pass through the
graph.  The RUP deletion proof for input clause $C$ effectively proves that any
total assignment $\assign$ that satisfies the POG definition clauses
but does not satisfy $C$ will yield
$\assign(r) = 0$.  It starts with the set of literals
$\{ \obar{\lit} \mid \lit \in C\}$, describing the required condition for
assignment $\assign$ to falsify clause $C$.
It then
adds literals via unit propagation until a
conflict arises.    Unit literal $r$ gets
added right away, setting up a potential conflict.
